---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sparseprepr

<!-- badges: start -->
[![R-CMD-check](https://github.com/dmolitor/sparseprepr/workflows/R-CMD-check/badge.svg)](https://github.com/dmolitor/sparseprepr/actions)
[![Codecov test coverage](https://codecov.io/gh/dmolitor/sparseprepr/branch/main/graph/badge.svg?token=PCAC1RA7GE)](https://codecov.io/gh/dmolitor/sparseprepr?branch=main)
[![pkgdown](https://github.com/dmolitor/sparseprepr/workflows/pkgdown/badge.svg)](https://github.com/dmolitor/sparseprepr/actions)
<!-- badges: end -->

The goal of sparseprepr is to enable common pre-processing actions for sparse 
matrices and provide a more memory-efficient workflow for modeling at scale.

## Installation

Install from [Github](https://github.com) with:
```
# install.packages(devtools)
devtools::install_github("dmolitor/sparseprepr")
```

## Scope

sparseprepr functionality only supports sparse matrices coded in sorted 
compressed column-oriented form, formally of class `CsparseMatrix`. Although the
`Matrix` package also defines sorted compressed row-oriented form 
(`RsparseMatrix`) and triplet form (`TsparseMatrix`) sparse matrices, it makes
clear that "most operations with sparse matrices are performed using the 
compressed, column-oriented or CsparseMatrix representation," and that even when 
matrices are created in the `TsparseMatrix` or `RsparseMatrix` forms for 
convenience, "once it is created, however, the matrix is generally coerced to a 
CsparseMatrix for further operations."
```{r, include = FALSE}
x <- Matrix::rsparsematrix(10, 3, density = 0.9)
x <- cbind(x, x[, 3, drop = FALSE])
x[10, 3] <- x[10, 3] + 0.001
x <- cbind(x, 1, 0)
x <- cbind(x, x[, 2, drop = FALSE])
x <- cbind(x, c(rep(0, 9), 1))
colnames(x) <- c(
  paste0("x", 1:3),
  "cor_with_x3",
  "const_col",
  "const_col2",
  "dup_x2",
  "sparse_col"
)
```

## Core Functionality

The following toy example shows a number of the pre-processing features that
sparseprepr provides.
```{r}
library(sparseprepr)

x
```

The matrix shown above has a number of contrived features; 
column `cor_with_x3` is highly correlated with column `x3`, `const_col` and 
`const_col2` are zero-variance columns, `dup_x2` is identical to `x2`, 
and `sparse_col` is a highly sparse column. Common pre-processing steps provided
by sparseprepr include:

- Dropping highly correlated columns
    ```{r}
    remove_correlated(x, threshold = 0.99)
    ```

- Dropping empty columns
    ```{r}
    remove_empty(x)
    ```

- Dropping constant columns
    ```{r}
    remove_constant(x)
    ```

- Dropping duplicated columns
    ```{r}
    remove_duplicate(x)
    ```

- Dropping highly sparse columns
    ```{r}
    remove_sparse(x, threshold = 0.9)
    ```

- Transforming matrix columns:
    ```{r}
    # Append squared terms
    transform_cols(
      x[, 1:3], 
      fun = function(i) i^2, 
      which.cols = paste0("x", 2:3), 
      name.sep = "squared"
    )
    
    # Append cubed terms
    transform_cols(
      x[, 1:3], 
      fun = function(i) i^3, 
      which.cols = paste0("x", 2:3), 
      name.sep = "cubed"
    )
    ```

## Pipe Workflow

These same pre-processing steps can be utilized in a more user-friendly manner
via the magrittr pipe (`%>%`) or the base pipe (`|>` - R 4.1 or greater).
```{r}
x |>
  remove_constant() |>
  remove_correlated(threshold = 0.99) |>
  remove_duplicate() |>
  remove_sparse(threshold = 0.9) |>
  transform_cols(
    fun = function(i) i ^ 2,
    which.cols = paste0("x", 2:3),
    name.sep = "squared"
  ) |>
  transform_cols(
    fun = function(i) i ^ 3,
    which.cols = paste0("x", 2:3),
    name.sep = "cubed"
  )
```
